Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-29 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Jed edits Ch5 through 7

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90~mm by 38~mm.  A partial ceiling with a width of at least 1.2~m, measured from the back wall of the compartment was installed across the width of the compartment, this allowed for the plume to impact the ceiling and form a ceiling jet along the back wall, before the fire gases flowed up and out of the compartment and into the 6~m by 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7~mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2~m wide and 2.4~m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway. These experiments were conducted without instrumentation in the walls for several reasons:
\begin{enumerate}
\item to compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation
\item to provide more realistic conditions
\item to conduct additional replicate experiments
\end{enumerate}

A photograph of the compartment fire experimental arrangement is shown in Figure~\ref{Compartment_Test}. In the instrumented gypsum wallboard experiments, as the burners were moved away from the position against the wall, the fire effects on the wall diminished significantly.  Therefore the fire pattern compartment experiments were conducted with the burners against the wall or in the corner of the compartment.

@@

\begin{figure}[ht!]
\includegraphics[width=\columnwidth]{../Figures/Compartment_Test} \\
	\caption[Photograph of the wall fire pattern experiments compartment arrangement]{Photograph of the wall fire pattern experiments compartment arrangement.  Experiment shown is being conducted with a polyurethane burner against the back wall.}
	\label{Compartment_Test}
\end{figure}

@@

 The initial set of compartment experiments were conducted to examine if the type of wall construction would impact the development of the fire patterns. These experiments determined how the walls for the remaining compartment fire pattern experiments were constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum wallboard on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3) Insulated, 12.7~mm thick, regular core, painted gypsum wallboard on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting, with a kraft paper facing, and an ``R value'' of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum wallboard. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

Ten replicate experiments were conducted with each of the three wall construction types.  The walls were exposed to both the natural gas and the gasoline burners.  For all of these experiments the burner was in contact with the gypsum wallboard.  A total of 60 experiments were conducted.  The results will be presented by wall construction type.  The order of presentation will be open, closed, and insulated wall construction.

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA:921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell or approximately 0.001~m$^2$.

There were four measures that were taken to characterize each pattern: maximum height, maximum width, height at maximum width, and the area.  Each dimensional value is presented with the 95\% confidence level (2~$\sigma$) based on Type A statistical analysis of the fire pattern measurements.

There are also a number of qualitative terms from NFPA~921 that were used to describe features of the patterns.  For example, ``clean burn'' to describe a lack of soot within a fire pattern area or ``inverted cone'' to describe the shape of the pattern~\cite{NFPA:921}.

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized in an effect to make the images comparable.
The average maximum height of the fire pattern generated by the natural gas burner on the open walls was 0.74~m~$\pm~0.12\%$.  The average maximum width of the fire pattern from the same experiments was 0.24~m~$\pm~0.06\%$.  The position from the top edge of the burner to the vertical position of the maximum width was 0.41~m~$\pm~0.07\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.05\%$.

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each patterns is similar to the width of the burner. Moving up from the burner, about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and burns, the plume and the pattern expands.  As the flames move up the wall the width tends to increase until about half of the peak height of the pattern, then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame is reduced.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticeable bias to one side or the other can be seen in four of the ten experiments.

The fire pattern photographs from the ten closed wall construction experiments are shown in Figure~\ref{NG_Closed_Wall}. 
The patterns appear very similar in nature and size to the patterns that were generated on the open construction walls.  The average maximum height of the fire patterns generated by the natural gas burner on the closed construction wall was the same as the patterns generated on the open construction wall, 0.74~m~$\pm~0.06\%$.  The 95~$\%$ confidence limit is about half of the value from the open wall experiments. The average maximum width of the fire patterns was 0.24~m~$\pm~0.02\%$.  Again similar to the previous experiments. The vertical position of the maximum width was 0.43~m~$\pm~0.03\%$. The area of the burn estimated by examining the damaged grid cells was 0.14~$m^2~\pm~0.01\%$.

@@

	\caption[Photographs of the fire patterns from ten replicate natural gas burner against open wall construction]{Photographs of the fire patterns from ten replicate natural gas burner against open wall construction.}
	\label{NG_Open_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from ten replicate natural gas burner against closed wall construction]{Photographs of the fire patterns from ten replicate natural gas burner against closed wall construction.}
	\label{NG_Closed_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from ten replicate natural gas burner against insulated wall construction]{Photographs of the fire patterns from ten replicate natural gas burner against insulated wall construction.}
	\label{NG_Insulated_Wall}
\end{figure}

@@

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each patterns is similar to the width of the burner. Moving up from the burner, about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and burns, the plume and the pattern expands.  As the flames move up the wall the width tends to increase until about half of the peak height of the pattern, then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame is reduced.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticeable bias to one side or the other can be seen in four of the ten experiments.

The fire pattern photographs from the ten closed wall construction experiments are shown in Figure~\ref{Gas_Closed_Wall}. 
The patterns appear very similar in nature and size to the patterns that were generated on the open construction walls.  The average maximum height of the fire patterns generated by the natural gas burner on the closed construction wall was the same as the patterns generated on the open construction wall, 0.74~m~$\pm~0.06\%$.  The 95~$\%$ confidence limit is about half of the value from the open wall experiments. The average maximum width of the fire patterns was 0.24~m~$\pm~0.02\%$.  Again similar to the previous experiments. The vertical position of the maximum width was 0.43~m~$\pm~0.03\%$. The area of the burn estimated by examining the damaged grid cells was 0.14~$m^2~\pm~0.01\%$.

The basic shape of these patterns on the closed wall are very similar to the patterns on the open wall.  Six of the ten patterns have a bump out or an unsymmetrical curve within 200~mm of the top of the burner.  The exact cause of these bumps in the patterns could not be determined, but they were generated later in the burn after the most of the fire pattern was created. Three of the patterns have a noticeable lean.

The last set of wall experiments with the natural gas burner was conducted with the insulated wall.  The patterns from the ten replicate experiments are shown in Figure~\ref{Gas_Insulated_Wall}.  The average maximum height of the fire patterns generated by the natural gas burner on the insulated walls was 0.71~m~$\pm~0.10\%$.  The average maximum width of the fire patterns on the insulated walls were 0.23~m~$\pm~0.04\%$.  The vertical position of the maximum width was 0.42~m~$\pm~0.07\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.02\%$.

@@

	\caption[Photographs of the fire patterns from ten replicate gasoline burner against open wall construction]{Photographs of the fire patterns from ten replicate gasoline burner against open wall construction.}
	\label{Gas_Open_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from ten replicate gasoline burner against closed wall construction]{Photographs of the fire patterns from ten replicate gasoline burner against closed wall construction.}
	\label{Gas_Closed_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from ten replicate gasoline burner against insulated wall construction]{Photographs of the fire patterns from ten replicate gasoline burner against insulated wall construction.}
	\label{Gas_Insulated_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from ten replicate polyurethane foam burner against open wall construction]{Photographs of the fire patterns from ten replicate polyurethane foam burner against open wall construction.}
	\label{PUF_Open_Wall}
\end{figure}

@@

Compartment experiments with the burner positioned in or near the corner were conducted with the open back wall construction.  The purpose of moving the burner position toward the corner was to examine the effect of reducing the available air entrainment area around the perimeter of the burner on the resulting fire pattern.  With the burner tight against both walls, the air entrainment perimeter was reduced by 50\%. Previous studies on flame height with fires in corners have demonstrated the the flame height increased when compared to a similar fire positioned against a single wall or burning in the open~\cite{Poreh:1999,Saito:1993,Takahashi:1997,Kokkala:1993}.  The previous studies measured temperatures, heat flux, and mass flow rates for the purpose of developing correlations to predict flame height. The sole purpose of these experiments was to examine the fire patterns that result from well characterized fires on painted gypsum wall board.

@@

The compartment used for the corner position experiments was the same as the compartment used for the wall positions experiments detailed in the previous section.  The burner was positioned in the left, rear corner against the face of two pieces of painted gypsum wallboard. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for complete extinguishment (total burn time) for the 500~mL of gasoline and the polyurethane urethane block. Typically the gasoline and polyurethane burners were in decay approximately 120~seconds after ignition. By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.  Figure~\ref{GBNG34Sequence} displays a sequence of photographs taken during a natural gas burner corner experiment.  The photographs taken at a rate of five images per second show the significant movement of the regions of visible combustion.  The grid drawn on the corner extends to 1.5 m above the top of the burner.

@@

The corner experiments demonstrated the expected extension of the combustion zone, especially in the cases of the natural gas and gasoline fires.  In the extreme case the flames from the gasoline burner briefly impinged on the ceiling, which is approximately 2.4~m above the top of the burner. The shape of the fire pattern to the wall was also different then the patterns generated on the flat wall.  Much smaller areas of the wall had the painted paper charred, penetrated, and burned.  For each of the fuels there was significantly more area where it was difficult to separate the charred areas from the surface deposits and discoloration. The sections below include post test images of the fire patterns.  The dimensions of the corner fire patterns are given in Table~\ref{tab:Fire_Pattern_Dimensions_Corners}.

@@

Figure~\ref{NG_Corner} shows a comparison of the ten fire patterns generated by the natural gas burner in the corner position with walls of open back construction. In the corner burns there is an area in the corner where the paper is not burned off.  From the flame sequence shown in Figure~\ref{GBNG34Sequence} it can be seen that the flame in the persistent flame region is close to the walls of the corner, but moving up into the intermittent region, the flame burns away from the walls, hence less energy is transferred to the walls and perhaps a bit more soot condenses out on the cooler walls.

@@

	\caption[Photographs of the fire patterns from ten replicate natural gas burner experiments positioned in a corner]{Photographs of the fire patterns from ten replicate natural gas burner experiments positioned in a corner.}
	\label{NG_Corner}
\end{figure}

@@

Figure~\ref{Gas_Corner} shows a comparison of eight of the ten fire patterns generated by the gasoline fueled burner in the corner position with walls of open back construction. Due to a camera malfunction there were no comparable images for the other two experiments.  The patterns are darker than the patterns generated on the flat walls.  Given that the air to the combustion zone was constrained it follows that the combustion may have been less efficient than the flat wall experiments which resulted in the sootier patterns.  The combustion zone for the gasoline fires positioned in the corner stretched from the top of the burner up to the ceiling of the compartment.  Figure~\ref{GBGas10Sequence} shows the flame extension and similar to the natural gas burner fires shows how the flame in the intermittent zone appeared to move away from the corner.

@@

	\caption[Photographs of the fire patterns from eight replicate gasoline burner experiments positioned in a corner]{Photographs of the fire patterns from eight replicate gasoline burner experiments positioned in a corner with walls of open construction.}
	\label{Gas_Corner}
\end{figure}

@@

Figure~\ref{PUF_Corner} shows a comparison of three fire patterns generated by the polyurethane foam fueled burner in the corner position with walls of open back construction. Only three replicates were conducted due to the limits on the foam supply that was sourced from the same purchase.  The flame length from the polyurethane foam fueled burner was also longer compared to the flat wall experiments.  However the change was not as big compared to the higher heat release rate natural gas and gasoline fueled burners.  The resulting patterns, similar to the other corner burner patterns, were darker in color with less definition in the lines of demarcation.

@@

	\caption[Photographs of the fire patterns from three replicate polyurethane foam burner experiments positioned in a corner]{Photographs of the fire patterns from three replicate polyurethane foam fueled burner experiments positioned in a corner with walls of open construction.}
	\label{PUF_Corner}
\end{figure}

@@

After the experiments the gypsum wallboard panels were removed from the compartment for documentation and measurement of the fire patterns.  A comparison on a fire pattern from each type of burner is shown in Figure~\ref{PT_Corners}.  Another image of the gasoline generated corner fire pattern, Figure~\ref{C_PTGASClose}, is included to show the reader how viewing a fire pattern at different angles enabled a better sense of the area of the fire damage.

@@

	\caption[Photographs of the fire patterns from three different burner experiments positioned in a corner]{Photographs of the fire patterns from three different burner experiments positioned in a corner with walls of open construction.  The patterns were generated by the following fuels, from are from left to right, natural gas, gasoline, polyurethane foam.}
	\label{PT_Corners}
\end{figure}

@@

	\caption[Photographs of a gasoline generated corner fire pattern taken at an angle]{Photographs of a gasoline generated corner fire pattern taken at an angle in order to improve the visibility of the lines of demarcation.}
	\label{C_PTGASClose}
\end{figure}

In general, the experiments, where the the burners were positioned against the center of a flat wall, provided very repeatable results with the natural gas and gasoline fueled burners. Under these pre-flashover, short burn duration fire conditions it was clear that the wall construction did not have any impact on the resulting fire pattern.  Also comparing these patterns to those generated with the instrumented gypsum wallboard, it did not appear that the instrumentation had any impact that changed the fire pattern.  With an expanded uncertainty of 2~$\sigma$, the height of the fire patterns were within $\pm$ 18\% for both the natural gas and the gasoline fuel burner fires.  The uncertainties regarding the width and area of the patterns were higher up to $\pm$ 33\%.  While the natural gas and the gasoline are different fuels there was no significant difference in any of the measured pattern dimensions between the two as show in Table~\ref{tab:Fire_Pattern_Dimensions_Wall_Construction}.  The larger uncertainty in the areas of the pattern may be due to the slight movement of the flames back and forth across the wall.

@@

The gasoline fueled fires also had similar fire pattern heights and flame heights; however, the similarity was not as strong as the natural gas fires.  The transient nature of the fire was believed to have had an impact since there was less energy transferred to the gypsum wall board and the exposure time for portions of the wall board more distant from the fuel surface was less.

>>>>

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long x 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90~mm x 38~mm.  A partial ceiling with a width of at least 1.2~m (measured from the back wall of the compartment) was installed across the width of the compartment, which allowed for the plume to impact the ceiling and form a ceiling jet along the back wall before the fire gases flowed up and out of the compartment and into the 6~m x 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7~mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2~m wide and 2.4~m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway.

These experiments were conducted without instrumentation in the walls for several reasons:
\begin{enumerate}
\item To compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation.
\item To provide more realistic conditions.
\item To conduct additional replicate experiments.
\end{enumerate}

A photograph of the compartment fire experimental arrangement is shown in Figure~\ref{Compartment_Test}. In the instrumented gypsum wallboard experiments, as the burners were moved away from the wall, the fire effects on the wall diminished significantly.  Therefore the fire pattern compartment experiments were conducted with the burners against the wall or in the corner of the compartment.

@@

\begin{figure}[ht!]
\includegraphics[width=\columnwidth]{../Figures/Compartment_Test} \\
	\caption[Photograph of the wall fire pattern experiments compartment arrangement]{Photograph of the wall fire pattern experiments compartment arrangement.  Experiment shown was conducted with a polyurethane burner against the back wall.}
	\label{Compartment_Test}
\end{figure}

@@

 The initial set of compartment experiments were conducted to examine if the type of wall construction would impact the development of the fire patterns. These experiments determined how the walls for the remaining compartment fire pattern experiments were constructed.  Three cases were examined: 1) open back, 12.7~mm thick, regular core, painted gypsum wallboard on the front of a wood frame only; 2) closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall; and 3) insulated, 12.7~mm thick, regular core, painted gypsum wallboard on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall.

 The insulation was 90~mm thick fiber glass batting, with a kraft paper facing, and an R value of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels (approximately 80$^\circ$C), although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were conducted with only the natural gas and gasoline fueled burners because the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum wallboard. As in the fire source characterization experiments, the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size, and burn duration.  The natural gas had a 300~seconds burn time, which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

Ten replicate experiments were conducted with each of the three wall construction types.  The walls were exposed to both the natural gas and the gasoline burners.  For all of these experiments, the burner was in contact with the gypsum wallboard.  A total of 60 experiments were conducted.  The results will be presented by wall construction type.  The order of presentation will be open, closed, and insulated wall construction.

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA:921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell, or approximately 0.001~m$^2$.

Four measures were taken to characterize each pattern: maximum height, maximum width, height at maximum width, and the area.  Each dimensional value is presented with the 95\% confidence level (2~$\sigma$) based on Type A statistical analysis of the fire pattern measurements.

A number of qualitative terms from NFPA~921 were used to describe features of the patterns.  For example, ``clean burn'' describes a lack of soot within a fire pattern area, and ``inverted cone'' describes the shape of the pattern~\cite{NFPA:921}.

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the 10 fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sizes to make the images comparable.  The average maximum height of the fire pattern generated by the natural gas burner on the open walls was 0.74~m~$\pm~0.12\%$.  The average maximum width of the fire pattern from the same experiments was 0.24~m~$\pm~0.06\%$.  The position from the top edge of the burner to the vertical position of the maximum width was 0.41~m~$\pm~0.07\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.05\%$.

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each pattern is similar to the width of the burner. Moving up from the burner about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and burns, the plume and the pattern expands.  As the flames move up the wall, the width tends to increase until about half of the peak height of the pattern.  Then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame decreases.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticeable bias to one side or the other can be seen in four of the 10 experiments.

The fire pattern photographs from the ten closed wall construction experiments are shown in Figure~\ref{NG_Closed_Wall}. 
The patterns appear very similar in nature and size to the patterns that were generated on the open construction walls.  The average maximum height of the fire patterns generated by the natural gas burner on the closed construction wall was the same as the patterns generated on the open construction wall, 0.74~m~$\pm~0.06\%$.  The 95~$\%$ confidence limit is about half the value from the open wall experiments. The average maximum width of the fire patterns was 0.24~m~$\pm~0.02\%$.  Again, similar to the previous experiments. The vertical position of the maximum width was 0.43~m~$\pm~0.03\%$. The area of the burn estimated by examining the damaged grid cells was 0.14~$m^2~\pm~0.01\%$.

@@

	\caption[Photographs of the fire patterns from 10 replicate natural gas burner against open wall construction experiments.]{Photographs of the fire patterns from 10 replicate natural gas burner against open wall construction experiments.}
	\label{NG_Open_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from 10 replicate natural gas burner against closed wall construction experiments.]{Photographs of the fire patterns from 10 replicate natural gas burner against closed wall construction experiments.}
	\label{NG_Closed_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from 10 replicate natural gas burner against insulated wall construction experiments.]{Photographs of the fire patterns from 10 replicate natural gas burner against insulated wall construction experiments.}
	\label{NG_Insulated_Wall}
\end{figure}

@@

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each pattern is similar to the width of the burner. Moving up from the burner about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and burns, the plume and the pattern expands.  As the flames move up the wall, the width tends to increase until about half of the peak height of the pattern.  Then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame decreases.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticeable bias to one side or the other can be seen in four of the 10 experiments.

The fire pattern photographs from the 10 closed wall construction experiments are shown in Figure~\ref{Gas_Closed_Wall}. The patterns appear very similar in nature and size to the patterns that were generated on the open construction walls.  The average maximum height of the fire patterns generated by the natural gas burner on the closed construction wall was the same as the patterns generated on the open construction wall, 0.74~m~$\pm~0.06\%$.  The 95~$\%$ confidence limit is about half the value from the open wall experiments. The average maximum width of the fire patterns was 0.24~m~$\pm~0.02\%$.  Again, similar to the previous experiments. The vertical position of the maximum width was 0.43~m~$\pm~0.03\%$. The area of the burn estimated by examining the damaged grid cells was 0.14~m$^2~\pm~0.01\%$.

The basic shape of these patterns on the closed wall are very similar to the patterns on the open wall.  Six of the 10 patterns have a bump out or an unsymmetrical curve within 200~mm of the top of the burner.  The exact cause of these bumps in the patterns could not be determined, but they were generated later in the burn after the most of the fire pattern was created. Three of the patterns have a noticeable lean.

The last set of wall experiments with the natural gas burner was conducted with the insulated wall.  The patterns from the 10 replicate experiments are shown in Figure~\ref{Gas_Insulated_Wall}.  The average maximum height of the fire patterns generated by the natural gas burner on the insulated walls was 0.71~m~$\pm~0.10\%$.  The average maximum width of the fire patterns on the insulated walls were 0.23~m~$\pm~0.04\%$.  The vertical position of the maximum width was 0.42~m~$\pm~0.07\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.02\%$.

@@

	\caption[Photographs of the fire patterns from 10 replicate gasoline burner against open wall construction experiments.]{Photographs of the fire patterns from ten replicate gasoline burner against open wall construction experiments.}
	\label{Gas_Open_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from 10 replicate gasoline burner against closed wall construction experiments.]{Photographs of the fire patterns from 10 replicate gasoline burner against closed wall construction experiments.}
	\label{Gas_Closed_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from 10 replicate gasoline burner against insulated wall construction experiments.]{Photographs of the fire patterns from 10 replicate gasoline burner against insulated wall construction experiments.}
	\label{Gas_Insulated_Wall}
\end{figure}

@@

	\caption[Photographs of the fire patterns from 10 replicate polyurethane foam burner against open wall construction experiments.]{Photographs of the fire patterns from 10 replicate polyurethane foam burner against open wall construction experiments.}
	\label{PUF_Open_Wall}
\end{figure}

@@

Compartment experiments with the burner positioned in or near the corner were conducted with the open back wall construction.  The purpose of moving the burner position toward the corner was to examine the effect of reducing the available air entrainment area around the perimeter of the burner on the resulting fire pattern.  With the burner tight against both walls, the air entrainment perimeter was reduced by 50\%. Previous studies on flame height with fires in corners have demonstrated the flame height increased when compared to a similar fire positioned against a single wall or burning in the open~\cite{Poreh:1999,Saito:1993,Takahashi:1997,Kokkala:1993}.  The previous studies measured temperatures, heat flux, and mass flow rates to develop correlations to predict flame height. The sole purpose of these experiments was to examine the fire patterns that result from well-characterized fires on painted gypsum wallboard.

@@

The compartment used for the corner position experiments was the same as the compartment used for the wall positions experiments detailed in the previous section.  The burner was positioned in the left, rear corner against the face of two pieces of painted gypsum wallboard. As in the fire source characterization experiments, the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~second burn time, which was approximately the upper bound for complete extinguishment (i.e. total burn time) for the 500~mL of gasoline and the polyurethane urethane block. Typically the gasoline and polyurethane burners were in decay approximately 120~seconds after ignition. By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.  Figure~\ref{GBNG34Sequence} displays a sequence of photographs taken during a natural gas burner corner experiment.  The photographs taken at a rate of five images per second show the significant movement of the regions of visible combustion.  The grid drawn on the corner extends to 1.5 m above the top of the burner.

@@

The corner experiments demonstrated the expected extension of the combustion zone, especially in the cases of the natural gas and gasoline fires.  In an extreme case, the flames from the gasoline burner briefly impinged on the ceiling, which was approximately 2.4~m above the top of the burner. The shape of the fire pattern to the wall was also different than the patterns generated on the flat wall.  Much smaller areas of the wall had the painted paper charred, penetrated, and burned.  For each of the fuels, there was significantly more area where it was difficult to separate the charred areas from the surface deposits and discoloration. The sections below include post test images of the fire patterns.  The dimensions of the corner fire patterns are given in Table~\ref{tab:Fire_Pattern_Dimensions_Corners}.

@@

Figure~\ref{NG_Corner} shows a comparison of the 10 fire patterns generated by the natural gas burner in the corner position with walls of open back construction. In the corner burns there is an area in the corner where the paper is not burned off.  The flame sequence in Figure~\ref{GBNG34Sequence} shows the flame in the persistent flame region is close to the walls of the corner, but moving up into the intermittent region, the flame burns away from the walls.  Hence, less energy is transferred to the walls, and perhaps a bit more soot condenses out on the cooler walls.

@@

	\caption[Photographs of the fire patterns from 10 replicate natural gas burner experiments positioned in a corner.]{Photographs of the fire patterns from 10 replicate natural gas burner experiments positioned in a corner.}
	\label{NG_Corner}
\end{figure}

@@

Figure~\ref{Gas_Corner} shows a comparison of eight of the 10 fire patterns generated by the gasoline fueled burner in the corner position with walls of open back construction. Due to a camera malfunction, there were no comparable images for the other two experiments.  The patterns are darker than the patterns generated on the flat walls.  Given that the air to the combustion zone was constrained, it follows that the combustion may have been less efficient than the flat wall experiments, which resulted in the sootier patterns.  The combustion zone for the gasoline fires positioned in the corner stretched from the top of the burner up to the ceiling of the compartment.  Figure~\ref{GBGas10Sequence} shows the flame extension and, similar to the natural gas burner fires, shows how the flame in the intermittent zone appeared to move away from the corner.

@@

	\caption[Photographs of the fire patterns from eight replicate gasoline burner experiments positioned in a corner.]{Photographs of the fire patterns from eight replicate gasoline burner experiments positioned in a corner with walls of open construction.}
	\label{Gas_Corner}
\end{figure}

@@

Figure~\ref{PUF_Corner} shows a comparison of three fire patterns generated by the polyurethane foam fueled burner in the corner position with walls of open back construction. Only three replicates were conducted due to the limits of the foam supply that was sourced from the same purchase.  The flame length from the polyurethane foam fueled burner was also longer compared to the flat wall experiments.  However, the change was not as big compared to the higher heat release rate of the natural gas and gasoline fueled burners.  The resulting patterns, similar to the other corner burner patterns, were darker in color with less definition in the lines of demarcation.

@@

	\caption[Photographs of the fire patterns from three replicate polyurethane foam burner experiments positioned in a corner.]{Photographs of the fire patterns from three replicate polyurethane foam fueled burner experiments positioned in a corner with walls of open construction.}
	\label{PUF_Corner}
\end{figure}

@@

After the experiments, the gypsum wallboard panels were removed from the compartment for documentation and measurement of the fire patterns.  A comparison on a fire pattern from each type of burner is shown in Figure~\ref{PT_Corners}.  Another image of the gasoline generated corner fire pattern, Figure~\ref{C_PTGASClose}, is included to show  how viewing a fire pattern at different angles enables a better sense of the area of the fire damage.

@@

	\caption[Photographs of the fire patterns from three different burner experiments positioned in a corner.]{Photographs of the fire patterns from three different burner experiments positioned in a corner with walls of open construction.  The patterns were generated by the following fuels from are from left to right: natural gas, gasoline, polyurethane foam.}
	\label{PT_Corners}
\end{figure}

@@

	\caption[Photographs of a gasoline generated corner fire pattern taken at an angle.]{Photographs of a gasoline generated corner fire pattern taken at an angle to improve the visibility of the lines of demarcation.}
	\label{C_PTGASClose}
\end{figure}

In general, the experiments in which the burners were positioned against the center of a flat wall provided very repeatable results with the natural gas and gasoline fueled burners. Under these pre-flashover, short burn duration fire conditions, it was clear the wall construction did not have any impact on the resulting fire pattern.  And when comparing these patterns to those generated with the instrumented gypsum wallboard, it did not appear that the instrumentation had any impact that changed the fire pattern either.  With an expanded uncertainty of 2~$\sigma$, the height of the fire patterns were within $\pm$ 18\% for both the natural gas and the gasoline fuel burner fires.  The uncertainties regarding the width and area of the patterns were higher up to $\pm$ 33\%.  Although the natural gas and the gasoline are different fuels, there was no significant difference in any of the measured pattern dimensions between the two as show in Table~\ref{tab:Fire_Pattern_Dimensions_Wall_Construction}.  The larger uncertainty in the areas of the pattern may be due to the slight movement of the flames back and forth across the wall.

@@

The gasoline fueled fires also had similar fire pattern heights and flame heights; however, the similarity was not as strong as the natural gas fires.  The transient nature of the fire was believed to have had an impact because there was less energy transferred to the gypsum wallboard, and the exposure time for portions of the wallboard more distant from the fuel surface was less.
